---
title: "01_rna_celltyping.Rmd"
author: "tobiasmohl"
date: "2025-07-24"
output: html_document
---
# 1. Setup
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

suppressPackageStartupMessages({
library(Seurat)
library(scRNAseq)
library(Signac)
library(GenomeInfoDb)
library(Matrix)
library(ggplot2)
library(SingleCellExperiment)
library(data.table)
library(pheatmap)
library(viridis)
library(scran)
library(edgeR)
library(scran)
library(leidenAlg)
library(HDF5Array) 
library(ensembldb)
library(EnsDb.Hsapiens.v86) 
library(scater) 
library(scDblFinder) 
library(batchelor)
library(matrixStats)
library(harmony)
library(tools)
library(sechm)
library(curl)
library(AnnotationHub)
library(AnnotationDbi)})

SCRIPT_DIR <- getwd()   
PROJECT_ROOT <- file.path(SCRIPT_DIR, "..")                     
RAW_DIR      <- file.path(PROJECT_ROOT, "data", "raw_data")
dir.create(RAW_DIR, recursive = TRUE, showWarnings = FALSE)
```

# 2. Download samples
```{r GSM sample mapping and download}
# GSM→Sample mapping 
# If a TSV exists, we read it; otherwise fall back to the embedded data.frame.

map_file <- file.path(PROJECT_ROOT, "data", "samples.tsv")
if (file.exists(map_file)) {
  samples <- fread(map_file)   # columns: GSM, sample
} else {
  samples <- data.table(
    GSM=c("GSM5818649","GSM5818650","GSM5818651","GSM5818652","GSM5818653","GSM5818654","GSM5818655","GSM5818656","GSM5818657",
          "GSM5818658","GSM5818659","GSM5818660","GSM5818661","GSM5818662","GSM5818663","GSM5818664","GSM5818665","GSM5818666",
          "GSM5818667","GSM5818668","GSM5818669","GSM5818670","GSM5818671","GSM5818672","GSM5818673","GSM5818674","GSM5818675",
          "GSM5818676","GSM5818677","GSM5818678","GSM5818679","GSM5818680","GSM5818681","GSM5818682"),
    sample=c("C16007","C4176","C4322","C4361","C4634","C4660","C4691","C4814","C5410",
             "P02284","P4205","P4560","P4653","P4772","P4884","P4919","P5318","P5331",
             "P5626","P5662","P6259","P6320","P6323","P6326","YC4345","YC5276","YC5346",
             "YC5442","YC5626","YC5751","YC5813","YC5928","YC6235","YC914")
  )
} 

suffixes <- c("filtered_feature_bc_matrix.h5", "atac_fragments.tsv.gz", "atac_peaks.bed.gz")
geo_url  <- function(gsm, sample, suffix){
  sprintf("https://www.ncbi.nlm.nih.gov/geo/download/?acc=%s&format=file&file=%s_%s_%s",
          gsm, gsm, sample, suffix)
}

download_one <- function(url, dest, retries=3){
  for (i in seq_len(retries)){
    ok <- tryCatch({
      curl::curl_download(url, destfile = dest, quiet = FALSE)
      file.exists(dest) && file.info(dest)$size > 0
    }, error = function(e) FALSE)
    if (ok) return(TRUE)
    if (file.exists(dest)) unlink(dest)
    Sys.sleep(2 * i)
  }
  FALSE
}

# Init manifest + counter 
manifest <- vector("list", nrow(samples) * length(suffixes))
k <- 0L

# --- Download loop ---
for (i in seq_len(nrow(samples))){
  gsm    <- samples$GSM[i]
  sample <- samples$sample[i]
  outdir <- file.path(RAW_DIR, gsm)
  dir.create(outdir, showWarnings = FALSE, recursive = TRUE)

  for (sfx in suffixes){
    fn   <- sprintf("%s_%s_%s", gsm, sample, sfx)
    url  <- geo_url(gsm, sample, sfx)
    dest <- file.path(outdir, fn)

    message(sprintf("→ %s", fn))
    if (!download_one(url, dest)) {
      warning(sprintf("Missing/empty: %s", fn))
      next
    }

    k <- k + 1L
    manifest[[k]] <- data.table(
      GSM=gsm, sample=sample, file=fn, url=url,
      bytes=file.info(dest)$size,
      md5=unname(md5sum(dest))
    )
  }
}

# log
if (k > 0L){
  manifest_dt <- rbindlist(manifest[seq_len(k)], fill=TRUE)
  man_path <- file.path(RAW_DIR, "download_manifest.tsv")
  fwrite(manifest_dt, man_path, sep="\t")
  message(sprintf("Done. %d files saved. Manifest: %s", k, man_path))
} else {
  message("No files were downloaded.")
}
```

# 3. Combine samples
## Subset to protein coding genes
```{r}
scefiles <- list.files(RAW_DIR, pattern = "filtered_feature_bc_matrix\\.h5$", recursive = TRUE, full.names = TRUE)
sample_names <- basename(dirname(scefiles))  

ensdb <- EnsDb.Hsapiens.v86::EnsDb.Hsapiens.v86
g <- genes(ensdb)
protein_coding_genes <- g$gene_id[g$gene_biotype == "protein_coding"]

tenx_list <- lapply(seq_along(scefiles), function(i) {
  m <- TENxMatrix(scefiles[[i]], group = "matrix")  
  m <- m[rownames(m) %in% protein_coding_genes, ]
  colnames(m) <- paste(sample_names[i], colnames(m), sep = "_")
  as.matrix(m)  # load to memory
})
cm_combined <- do.call(cbind, tenx_list)
sce <- SingleCellExperiment(assays = list(rna_counts = cm_combined))
sce$Sample <- sub("_.*", "", colnames(sce))


assayNames(sce, "rna_counts") <- "counts"

sce_path <- file.path(PROJECT_ROOT, "data", "sce_full_proteincoding.rds")
saveRDS(sce, file = sce_path)
```

# 4. Gene Annotation
```{r}
sce$Sample <- factor(sce$Sample)

ensdb <- EnsDb.Hsapiens.v86::EnsDb.Hsapiens.v86
ens_ids <- rownames(sce)

rowData(sce)$ENSEMBL     <- ens_ids
rowData(sce)$SYMBOL      <- mapIds(ensdb, keys = ens_ids,
                                   keytype = "GENEID", column = "SYMBOL")
rowData(sce)$SEQNAME     <- mapIds(ensdb, keys = ens_ids,
                                   keytype = "GENEID", column = "SEQNAME")
rowData(sce)$GENEBIOTYPE <- mapIds(ensdb, keys = ens_ids,
                                   keytype = "GENEID", column = "GENEBIOTYPE")

sym_safe <- ifelse(is.na(rowData(sce)$SYMBOL) | rowData(sce)$SYMBOL == "",
                   rowData(sce)$ENSEMBL, rowData(sce)$SYMBOL)
rownames(sce) <- uniquifyFeatureNames(rowData(sce)$ENSEMBL, sym_safe)
```

# 5. Quality Control
## Low-Quality cells
```{r}
# identifying the mitochondrial transcripts
mito_idx <- which(rowData(sce)$SEQNAME %in% c("MT","chrM","M","chrMT"))
stats <- perCellQCMetrics(sce, subsets = list(Mt = mito_idx))
batch <- factor(colData(sce)$Sample)

qc <- quickPerCellQC(stats, percent_subsets = "subsets_Mt_percent", batch = batch)
reasons <- perCellQCFilters(stats, sub.fields = "subsets_Mt_percent", batch = batch)
colSums(as.matrix(reasons))
summary(reasons$discard)
attr(reasons$low_lib_size, "thresholds")
attr(reasons$low_n_features, "thresholds")

# Totals & barplot BEFORE filtering
n_total      <- length(qc$discard)
n_discarded  <- sum(qc$discard)
counts <- c(Kept = n_total - n_discarded, Discarded = n_discarded)
barplot(counts, col = c("darkgreen","tomato"),
        ylab = "Number of cells",
        main = "Overall QC Filtering (Kept vs Discarded)")

```
## Doublet Detection
```{r}
sce <- logNormCounts(sce)

sce <- scDblFinder(sce, clusters = TRUE, samples = sce$Sample)

sce <- sce[, sce$scDblFinder.class != "doublet" & sce$scDblFinder.score < 0.5]

sce_path <- file.path(PROJECT_ROOT, "data", "sce_filtered_proteincoding.rds")
saveRDS(sce, file = sce_path)
```

# 6. Celltyping
## Marker Genes
### Genes - Pierre Luc
```{r}
markerGenes <- list(
  astrocytes = c("Aqp4", "Gfap", "Fgfr3","Dio2","Acsbg1","Sox9","Ndrg2","Aldh1l1","Slc1a2","Slc1a3"),
  smooth_muscle = c("Tagln", "Cnn1", "Acta2", "Myl9", "Myh11"),
  pericytes = c("Aspn", "Notch3", "Ndufa4l2", "Pdgfrb", "Higd1b"),  
  endothelial = c("Cdh5","Flt1","Pecam1","Flt4","Abcg2","Vwf","Slc2a1","Bsg"),
  microglia = c("C1qb","Tyrobp","P2ry12", "Csf1r", "Irf8"),
  oligodendrocyte = c("Opalin","Plp1","Mag","Mog","Klk6","Mbp","Nipal4"),
  OPC = c("Pdgfra","Sox6","Bcan","Neu4", "Fyn", "Tnr"),
  COP = c("Fyn","Ust","Opcml","Tnr","Pcdh7","Enpp6"),
  mural = c("Slc6a20a","Vtn","Colec12", "Arhgap29", "Zic1"),
  PVM = c("Mrc1","F13a1","Dab2","Lyz2"),
  other_vascular = c("Bnc2","Mgp"),
  CR = c("Reln","Trp73"),
  neuronal = c("Snap25", "Stmn2", "Syn1", "Rbfox3", "Dlg4", "Gabbr2","Grin2b","Grin1"),
  neuronal_excitatory = c("Slc17a7","Slc17a6","Camk2a","Meis2","Neurod6","Nrn1"),
  neuronal_inhibitory = c("Gad1","Gad2","Lhx6","Adarb2","Slc6a1","Pvalb","Sst","Sema3c","Slc32a1","Npy"),
  neuroblast=c("Eomes", "Hopx", "Pou3f4","Ascl1", "Fabp7", "Nes", "Mki67"),
  #cycling=c("Top2a","Cdc25c"),
  others=c("Lamp5","Vip","Oxt","Oxtr","Chodl","Pax6","Sncg","Vipr2","Ntng1","Pax6","Nxph1","Ptprm")
)

geneMeta <- data.table(Symbol=rownames(rowData(sce)),
                       ID=as.numeric(1:nrow(rowData(sce))))

markerGenesTable <- data.table(Symbol=unlist(markerGenes, use.names=F),
                           Type=rep(names(markerGenes), lengths(markerGenes)))
markerGenesTable <- merge(markerGenesTable, geneMeta, by="Symbol", all.x=TRUE, suffixes="")
```

### Genes - Paper
```{r}
markerGenes <- list(
  Neurons        = c("STMN2","TBR1"),
  Oligodendrocytes = c("MBP","MOG"),
  Astrocytes     = c("GFAP","AQP4"),
  Microglia      = c("GPR34","TREM2","C1QC"),
  OPC            = c("OLIG1","OLIG2"),
  Endothelial    = c("FLT1","VWF","PDGFRB"),
  Tcells         = c("CD3E","CD8A")
)
geneMeta <- data.table(Symbol=rownames(rowData(sce)),
                       ID=as.numeric(1:nrow(rowData(sce))))

markerGenesTable <- data.table(Symbol=unlist(markerGenes, use.names=F),
                           Type=rep(names(markerGenes), lengths(markerGenes)))
markerGenesTable <- merge(markerGenesTable, geneMeta, by="Symbol", all.x=TRUE, suffixes="")
```

## Seurat, Normalize, Scale, PCA
```{r}
seo <- as.Seurat(sce, counts = "counts", data = "logcounts")


seo$Sample <- sce$Sample

seo <- NormalizeData(seo)
seo <- FindVariableFeatures(seo)
seo <- ScaleData(seo)
seo <- RunPCA(seo)

seo_batch <- RunHarmony(seo, group.by.vars = "Sample", verbose = TRUE)

sce_path <- file.path(PROJECT_ROOT, "data", "seo_filtered_proteincoding_harmony.rds")
saveRDS(seo_batch, file = sce_path)
```

## Weigh PCA, Plot Clusters
```{r}
# extract PCA embeddings
pca <- Embeddings(seo_batch, "harmony")
qc_metric <- log10(seo_batch$nCount_originalexp + 1)

# compute weights: penalize PCs correlated with QC
wf <- 1 - (cor(pca[, 1:50], qc_metric))^2

# apply weights
pca_dense <- as.matrix(pca[, 1:50])
pca_weighted <- sweep(pca_dense, 2, wf, `*`)

# save as new reduction
seo_batch[["merged_harmony_PCA"]] <- CreateDimReducObject(
  embeddings = pca_weighted,
  key = "mpca_",
  assay = Assays(seo_batch)
)

# run UMAP on the weighted PCA
seo_batch <- RunUMAP(seo_batch, reduction = "merged_harmony_PCA", dims = 1:50, reduction.name = "merged_UMAP")

# find neighbors and clusters based on merged PCA
seo_batch <- FindNeighbors(seo_batch, reduction = "merged_harmony_PCA", dims = 1:50, graph.name = "merged_PCA_snn")
seo_batch <- FindClusters(seo_batch, resolution = 4.5, group.singletons = FALSE,
                    algorithm = 4, random.seed = 1, graph.name = "merged_PCA_snn")

# set cluster identities
Idents(seo_batch) <- seo_batch$seurat_clusters

# UMAP plot
DimPlot(seo_batch, reduction = "merged_UMAP", label = TRUE) +
  ggtitle("Clusters (QC-weighted PCA)")

```

```{r}
seo <- seo_batch
sce <- as.SingleCellExperiment(seo, assay = "originalexp")

# Align by cell names before copying anything
stopifnot(identical(sort(colnames(seo)), sort(colnames(sce))))
ord <- match(colnames(sce), colnames(seo))

# Transfer cluster and key metadata (aligned!)
sce$cluster         <- seo$seurat_clusters[ord]
colData(sce)$Sample <- seo$Sample[ord]
colData(sce)$nUMI   <- seo$nCount_originalexp[ord]
colData(sce)$nGene  <- seo$nFeature_originalexp[ord]
colData(sce)$doublet_class <- seo$scDblFinder.class[ord]

# Transfer reductions with alignment
emb_umap <- Embeddings(seo, "merged_UMAP")
emb_pca  <- Embeddings(seo, "merged_harmony_PCA")
emb_umap <- emb_umap[match(colnames(sce), rownames(emb_umap)), , drop=FALSE]
emb_pca  <- emb_pca [match(colnames(sce), rownames(emb_pca )), , drop=FALSE]

reducedDim(sce, "merged_UMAP")       <- emb_umap
reducedDim(sce, "merged_harmony_PCA")<- emb_pca
```

## Aggregate cells
```{r}
sce_agg <- scuttle::aggregateAcrossCells(
  sce,
  ids = sce$cluster,
  use.assay.type = "counts",
  statistics = "sum"
)

rowData(sce_agg)$SYMBOL <- rownames(sce_agg)
rowData(sce_agg)$SYMBOL_UPPER <- toupper(rowData(sce_agg)$SYMBOL)

# TMM -> CPM
dge <- DGEList(counts = SummarizedExperiment::assay(sce_agg, "counts"))
dge <- calcNormFactors(dge)
SummarizedExperiment::assays(sce_agg)$cpm <- cpm(dge, log = FALSE)

# Per-gene 0–1 scaling (avoid divide-by-zero)
den <- pmax(rowMaxs(SummarizedExperiment::assays(sce_agg)$cpm), .Machine$double.eps)
SummarizedExperiment::assays(sce_agg)$scaled_cpm <- sweep(SummarizedExperiment::assays(sce_agg)$cpm, 1, den, "/")
```

## Heatmap
```{r}
# 1) Prepare marker table (uppercase)
markerGenes_upper <- toupper(unlist(markerGenes, use.names = FALSE))
marker_df <- data.table(
  Symbol = markerGenes_upper,
  Type   = rep(names(markerGenes), lengths(markerGenes))
)
marker_df <- unique(marker_df, by = "Symbol")

# 2) Pick rows present in aggregated object
keep_idx <- which(rowData(sce_agg)$SYMBOL_UPPER %in% marker_df$Symbol)
if (length(keep_idx) == 0L) stop("None of the marker genes were found in sce_agg.")

# 3) Extract matrix (scaled CPM) and set readable rownames
stopifnot("scaled_cpm" %in% assayNames(sce_agg))
mat <- assays(sce_agg)$scaled_cpm[keep_idx, , drop = FALSE]
rownames(mat) <- rowData(sce_agg)$SYMBOL_UPPER[keep_idx]

# 4) Build row annotation aligned to mat
marker_aln <- marker_df[match(rownames(mat), marker_df$Symbol)]
annoRow <- data.frame(type = marker_aln$Type, row.names = rownames(mat))

# 5) Order rows by marker type for nicer display
ord <- order(annoRow$type)
mat_ord <- mat[ord, , drop = FALSE]
annoRow_ord <- annoRow[ord, , drop = FALSE]

# 6) Plot
pheatmap(
  mat_ord,
  annotation_row = annoRow_ord,
  color = viridis(30),
  cluster_rows = FALSE,
  cluster_cols = FALSE,
  show_colnames = TRUE,
  fontsize_row = 7,
  fontsize_col = 10,
  main = "Scaled CPM of Marker Genes by Cluster"
)
```

## Manual Annotations - Pierre Luc - Specific
```{r}
manual_annotation <- list(
  astrocytes = c("20", "29", "42"),
  #COP = c(),
  #CR = c(),
  endothelial = c("52"),
  microglia = c("18","22", "38", "40", "51"),
  mural = c("44"),
  neuroblast = c("43"),
  neuronal = c("34", "39", "46"),
  neuronal_excitatory = c(),
  neuronal_inhibitory = c("30"),
  oligodendrocyte = c("2","3","4","5","6","7","8", "9", "10", "11", "12", "13", "14", "15", "16", "17", "19", "21","23", "24", "25", "26", "27", "28", "32", "33", "35", "41"),
  OPC = c("1", "37", "50"),
  #other_vascular = c(),
  others = c("31", "36", "48"),  
  pericytes = c("45", "49"),
  PVM = c("47")
  #smooth_muscle = c()
)

# cell type mapping dataframe 
man_annot_df <- data.frame(
  cluster = unlist(manual_annotation),
  cell_type = rep(names(manual_annotation), sapply(manual_annotation, length)),
  stringsAsFactors = FALSE
)

# annotate SCE object 
sce$cell_type_manual <- sapply(sce$cluster, function(i) {
  hit <- man_annot_df[man_annot_df$cluster == i, "cell_type"]
  if (length(hit) > 0) hit[1] else NA
})

# transfer annotation to Seurat object 
seo$cell_type_manual <- sce$cell_type_manual  # same cell order

# save updated Seurat object
saveRDS(seo, file = "/mnt/germain/work/tomohl/pgBoost-HumanData/data/seo_all_samples_proteincoding_annotated.rds")
```

## Manual Annotations - Broad
```{r}
broad_annotation <- list(
  Ast = c("20", "29", "42"),                                   # astrocytes
  End = c("52"),                                               # endothelial
  MG  = c("18", "22", "38", "40", "51"),                       # microglia
  OD  = c("2","3","4","5","6","7","8", "9", "10", "11",
          "12", "13", "14", "15", "16", "17", "19", "21",
          "23", "24", "25", "26", "27", "28", "32", "33",
          "35", "41"),                                         # oligodendrocyte
  OPC = c("1", "37", "50"),                                    # OPC
  IN  = c("30"),                                               # inhibitory neurons
  Ext = c("34", "39", "46"),                                   # excitatory neurons
  Other = c("31", "36", "48", "43", "44", "45", "47", "49")    # mural, pericytes, neuroblast, others
)
# cell type mapping dataframe 
man_annot_df <- data.frame(
  cluster = unlist(broad_annotation),
  cell_type = rep(names(broad_annotation), sapply(broad_annotation, length)),
  stringsAsFactors = FALSE
)

# annotate SCE object 
sce$cell_type_broad <- sapply(sce$cluster, function(i) {
  hit <- man_annot_df[man_annot_df$cluster == i, "cell_type"]
  if (length(hit) > 0) hit[1] else NA
})

# transfer annotation to Seurat object
seo$cell_type_broad <- sce$cell_type_broad  # same cell order

# save updated Seurat object 
saveRDS(seo, file = "/mnt/germain/work/tomohl/pgBoost-HumanData/data/seo_all_samples_proteincoding_annotated_broad.rds")

DimPlot(seo, group.by = "cell_type_broad", reduction = "merged_UMAP", label = TRUE, repel = TRUE) +
  ggtitle("Broad Cell Type Annotation") +
  theme_minimal()
```

## Manual Annotations - Paper
```{r}
manual_annotation <- list(
  astrocytes = c("25", "46"),
  endothelial = c("58"),
  Microglia = c("1", "32", "60"),
  Neurons = c("47", "52", "53", "56"),
  Oligodendrocytes = c("3", "4", "5", "6", "7", "8", "9", "10", "11", "12", "13", "14", "15", "16", "17", "18", "19", "20", "21", "22", "23", "26", "27", "28", "29", "30", "31", "33", "34", "35", "36", "37", "38", "39", "40", "41", "42", "43", "44", "45","48", "49", "50", "51", "54", "55", "57"),
  OPC = c("2", "24", "61", "62"),
  Tcells = c("59"),
  Other = ("63")
)

# cell type mapping dataframe 
man_annot_df <- data.frame(
  cluster = unlist(manual_annotation),
  cell_type = rep(names(manual_annotation), sapply(manual_annotation, length)),
  stringsAsFactors = FALSE
)

# annotate SCE object 
sce$cell_type_manual <- sapply(sce$cluster, function(i) {
  hit <- man_annot_df[man_annot_df$cluster == i, "cell_type"]
  if (length(hit) > 0) hit[1] else NA
})

# transfer annotation to Seurat object 
seo$cell_type_manual <- sce$cell_type_manual  # same cell order

# save updated Seurat object
sce_path <- file.path(PROJECT_ROOT, "data", "seo_annotated_paper.rds")
saveRDS(seo, file = sce_path)
```





